### 2024-10-07 draft-2/draft-2.tex
miscellaneous cleanup and mention of void messages

\minor Message terminators divide the client and engine streams into
\term{messages}; that is, a message is a (possibly empty) subsequence of bytes
that does not contain a message terminator and every byte of the client and
engine streams is either part of a message or part of a message terminator.
The messages within the client stream are \term{client messages} and the
messages within the engine stream are \term{engine messages}.
\vskip-\parskip
\greybox{This precludes the proper use of some encodings for client−engine
communication, such as \hbox{\caps utf-16} or \hbox{\caps utf-32} (since in
\hbox{\caps utf-16} and \hbox{\caps utf-32}, the byte \hex{0a} appears in the
encodings of various scalar values, whereas in \hbox{\caps utf-8}, the byte
\hex{0a} only appears in the encoding for \uni{000a line feed}).}
\vskip-\parskip
\greybox{Note that \hex{00} is not disallowed.}

@@

@@

\minor {\caps todo}\hem a client message has one of seven types indicated
by the first token of the message (\s{debug}, \s{setoption}, \s{ucinewgame},
\s{position}, \s{isready}, \s{go}, and \s{stop})

@@

\minor {\caps todo}\hem an engine message has one of six types indicated
by the first token of the message (\s{id}, \s{option}, \s{uciok}, \s{info},
\s{readyok}, \s{bestmove})

@@

@@

\minor In some states, the engine may dequeue a message. If the engine dequeues
a message, the message must be well-formed and have an allowed message type
given the state of the engine.
\vskip-0.5\baselineskip
\greybox{This is not meant to imply that standard input is ever required to be
empty, but meant to indicate that in some states the engine is not allowed to
read standard input.}

@@

\minor The allowed message types that may be dequeued and enqueued in each state
are enumerated in the following table. If the {\caps next} column is not empty
for a particular row, the action listed in the row (dequeuing or enqueuing a
message of a particular type) causes a transition to the listed state.

@@

{\caps todo}\hem pondering

\minor {\caps todo}\hem clients are required to send \s{ucinewgame} at least
once before the very first time they send \s{go} (but engines are strongly
recommended to behave identically whether or not \s{ucinewgame} is ever sent
or not)

\minor {\caps todo}\hem clients are required to send \s{position} at least
once before the very first time they send \s{go} (but engines are strongly
recommended to behave as if \s{position startpos} were sent if \s{position}
is not sent before \s{go} is sent)

{\caps todo}\hem note that the client is not required to send \s{isready}
before \s{go}

{\caps todo}\hem note that \s{ucinewgame} does not reset the current board
position nor does it reset the move history (only \s{position} changes the
current board state and move history)

{\caps todo}\hem note that the client may send \s{ucinewgame} after \s{position}
before \s{go} without any additional intervening \s{position}

{\caps todo}\hem note that engines and clients are not required to ignore
or otherwise handle unknown or unexpected tokens (these are simply violations)

@@

@@

\vskip\parskip
\beginruby
format(%q{
def trimCR('seq : List(Byte)) : List(Byte)
  if empty?('seq) or last('seq) ≠ `0d then return 'seq
  return withoutLast('seq)
})
\endruby

\vskip\parskip
\beginruby
format(%q{
def read_message!('fd : File_Descriptor) : List(Byte) | EOF
  'seq <- empty
  repeat
    match read_byte!('fd)
      eof => return (\pt if empty?('seq) then eof else 'seq\pt)
      `0a => return trimCR('seq)
      'val => append!('seq, 'val\pt)
})
\endruby

\vskip\parskip
\beginruby
format(%q{
def get_token!('seq : List(Byte)) -> List(Byte) | None
  repeat
    if empty?('seq) then return none
    if first('seq) ≠ `20 then break
    removeFirst!('seq)
  'tok <- empty
  repeat
    append!('tok, first('seq))
    removeFirst!('seq)
    if empty?('seq) then return 'tok
    if first('seq) = `20 then break
  repeat
    removeFirst!('seq)
    if empty?('seq) then return 'tok
    if first('seq) ≠ `20 then break
  return 'tok
})
\endruby

>>>>

\minor Message terminators divide the client and engine streams into
\term{messages}; that is, a message is a (possibly empty) sequence of bytes
that does not contain a message terminator, and every byte of the client and
engine streams is either part of a message or part of a message terminator.
The messages within the client stream are \term{client messages} and the
messages within the engine stream are \term{engine messages}.
\vskip-\parskip
\greybox{This precludes the proper use of some encodings for client−engine
communication, such as \hbox{\caps utf-16} or \hbox{\caps utf-32} (since in
\hbox{\caps utf-16} and \hbox{\caps utf-32}, the byte \hex{0a} appears in the
encodings of various scalar values, whereas in \hbox{\caps utf-8}, the byte
\hex{0a} only appears in the encoding for \uni{000a line feed}).}
\vskip-\parskip
\greybox{Note that \hex{00} is not disallowed.}

@@

\minor A message that does not contain any tokens (that is, an empty sequence
or a sequence of one or more repetitions of \hex{20}) is \term{void}.

@@

A client message has one of eight types indicated by the first token of the
message (\s{uci}, \s{debug}, \s{setoption}, \s{ucinewgame}, \s{position},
\s{isready}, \s{go}, and \s{stop}).

@@

An engine message has one of six types indicated by the first token of the
message (\s{id}, \s{option}, \s{uciok}, \s{info}, \s{readyok}, \s{bestmove}).

@@

\minor The client and the engine both may always enqueue or dequeue a void
message. When the client or engine dequeues a void message, it must ignore
the message, behaving as if the message were not transmitted.

The remainder of section \the\majorcount\ is written without mention of void
messages; consideration for void messages is tacit.

% Empty messages may always be sent and must be ignored by the client
% and engine. Empty messages never cause a transition.

@@

\minor In some states, the engine may dequeue a message. If the engine dequeues
a message, the message must be well-formed and have an allowed message type
given the state of the engine.
\vskip-\parskip
\greybox{This is not meant to imply that standard input is ever required to be
empty, but meant to indicate that in some states the engine is not allowed to
read standard input (or at least must behave as if it has not read standard
input).}

@@

\minor The allowed message types that may be dequeued and enqueued in each state
are enumerated in the following table. If the {\caps next} column is not empty
for a row, the action listed in the row (dequeuing or enqueuing a message of
a particular type) causes a transition to the listed state.

@@

\vglue-3\baselineskip\vfil\break

\greybox{Engines and clients are not required to ignore (or otherwise handle)
unknown or unexpected tokens (these are simply violations).}

\minor The client must enqueue a \s{ucinewgame} message at least once before
before it enqueues a \s{go} message for the first time.
\vskip-\parskip
\bluebox{Engines are strongly recommended to behave identically whether or not
a \s{ucinewgame} message is ever sent.}

\minor The client must enqueue a \s{position} message at least once before it
enqueues a \s{go} message for the first time.
\vskip-\parskip
\bluebox{Engines are strongly recommended, if a \s{position} message is not
sent before the first \s{go} message, to behave as if \s{position startpos}
were sent.}
\vskip-\parskip
\greybox{Note that the client is not required to send \s{isready} before
\s{go}.}
\vskip-\parskip
\greybox{Note that the client may send \s{ucinewgame} after \s{position}
before \s{go} without sending any additional intervening \s{position} message.}
\vskip-\parskip
\greybox{Note that \s{ucinewgame} does not reset the current board state nor
does it reset the move history. (Only \s{position} changes the current board
state and move history.)}

% This specification stops governing the interaction between the client
% and the engine when the client sends a well-formed quit message in
% the idle state. Engines are recommended to shut down and terminate
% themselves. Clients are recommended to provide a grace period before
% killing the engine.

@@

\advance\hsize by 10mm

@@

\vskip\parskip
\beginruby
format(%q{
def withoutCR('msg : List(Byte)) : List(Byte)
  if empty?('msg) or last('msg) ≠ `0d then return 'msg
  return withoutLast('msg)
})
\endruby

\vskip\parskip
\beginruby
format(%q{
def read_message!(mut 'fd : File_Descriptor) : List(Byte) | EOF
  'msg <- empty
  repeat
    match read_byte!('fd)
      eof => return (\pt if empty?('msg) then eof else 'msg\pt)
      `0a => return withoutCR('msg)
      'val => append!('msg, 'val\pt)
})
\endruby

\vskip\parskip
\beginruby
format(%q{
def get_token!(mut 'msg : List(Byte)) : List(Byte) | None
  repeat
    if empty?('msg) then return none
    if first('msg) ≠ `20 then break
    removeFirst!('msg)
  'tok <- empty
  repeat
    append!('tok, first('msg))
    removeFirst!('msg)
    if empty?('msg) then return 'tok
    if first('msg) = `20 then break
  return 'tok
})
\endruby
